---
title: "CytoexploreR Vignette for Gresham Lab"
author: "Julie Chuong"
output: html_notebook
---

This is notebook is to show you how to use the package CytoexploreR to analyze flow cytometry data from the Cytek Aurora. 

Install CytoExplorer package and requirements (Skip if already installed)
```{r}
library(BiocManager)
install.packages("cytolib", "flowCore", "flowWorkspace", "openCyto")
```

Install CytoExploreR from GitHub (Skip is already installed)
```{r}
library(devtools)
devtools::install_github("DillonHammill/CytoExploreR")
```

Load required packages
```{r, include = FALSE, echo=FALSE}
library(CytoExploreR)
library(tidyverse)
library(ggridges)
```

## Vignette Dataset

Three replicate populations of the GAP1 CNV reporter strain, in which a mCitrine gene
was inserted upstream of the *GAP1* promoter and coding sequence, were experimentally evolved in
glutamine-limited media in chemostats for 200 generations. Additionally, there was one population each of
3 control strains: zero copy 'unstained' control (DGY1) that has no mCitrine, one copy control (DGY500) in
which the mCitrine gene was inserted in neutral locus *WHAT LOCUS?* that doesn't undergo copy number
variation under glutamine-limited growth, and two copy control (DGY1315) containing two inserted copies of
mCitrine in neutral loci. Samples of each population were taken every 8, 13, 8 generations (Mon, Wed, Fri). 
In each sample, 100,000 events were measured on the Cytek Aurora Flow Cytometer for mCitrine fluorescence (excitation
wavelength = 516, emission wavelength = 529 ) via the B2-A channel, forward scatter (FSC), and side scatter (SSC). 
In total, there are six populations and 19 timepoints. 

## Set working directory and get list of subdirectories
Place your FCS files in a directory. Multiple subdirectories containing FCS files are OK. 

*Note 1* You can only load in one folder's worth of FCS files at a time. In this vignette, 
we have organized one timepoint's worth of FCS files per subdirectory. There is one FSC file per population.
Additionally, we have one subdirectory with three timepoints' worth of FCS files inside for drawing gates (because
timepoint one data were not sufficient to draw good gates, see STEP __). 
In total, there are 20 subdirectories, with an FSC file per population (for which there are six).

*Note 2* The subdirectories are uniquely named like this TimepointNumber_ExperimentName_SampleDate_generation_Initials.
Therefore our subdirectory names contain important  metadata. 

*Note 3* You should not manually rename your FCS files after exporting it from the Cytek. 
In other words, files should have their original names from the machine. 
"Renaming the files manually is not recommended as it does not update the associated keywords contained in the FCS files"
(Source: https://github.com/DillonHammill/CytoExploreR/issues/85). If files need to renamed, do so on on the Cytek
software and re-export the FCS files.  

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
knitr::opts_knit$set(root.dir = "/Volumes/GoogleDrive/My Drive/greshamlab/projects/EE_GAP1_ArchMuts_Summer2021/data/vignette_data_workingcopy")
```

```{r}
getwd()
folders = list.dirs()[-1]

```

## Version name
Choose a name to be used for all output files including the gating template and associated flow data and graphs.
In this vignette, since we will be using data from timepoints 1, 2, and 4 to draw gates, we choose a name to reflect that. 
```{r}
version_name = "timepoints-01-02-04_v1"
getwd()
```

## STEP 1: Generate experiment details file.
A .csv file that contains the list of .fcs files in the directory and the associated metadata for each sample.

A custom sample sheet is imported to the function. 
This function uses regex (regular expressions) to extract metadata from the subdirectory name and sample sheet 
to generate an experiment details file for each timepoint subdirectory and places the file inside the corresponding
subdirectory. We sampled cells from a 96 well plate so our metadata contains well number. 
```{r}
make_exp_details = function(folder_name, samplesheet) {
  pref = folder_name %>% str_extract("([0-9])+_EE_GAP1_ArchMuts_2021")
  generation = folder_name %>% str_extract("[g]\\d+") %>% str_remove("g")

  files = as_tibble(list.files(paste0(folder_name))) %>%
    separate(value, into = c("well", "samp"), sep = " ", remove = F) %>%
    mutate(well = str_extract(well, "([A-Z])([0-9]){1,2}$")) %>%
    mutate(samp = str_remove(samp, ".fcs")) %>%
    mutate(sample = case_when(str_detect(value, "Unstained") ~ "ctrl0",
                              str_detect(value, "DGY500") ~ "ctrl1",
                              str_detect(value, "DGY1315") ~ "ctrl2",
                              TRUE ~ samp)) %>%
    select(value,sample) %>%
    rename(name = value) %>%
    filter(!is.na(sample))

  all = files %>%
    left_join(read_csv(paste0("./",samplesheet)), by = c("sample" = "Sample name")) %>%
    mutate(generation = as.numeric(generation))

  write_csv(all, file = paste0(folder_name,"/",pref,"_experiment_details.csv"))

}
```

Only needs to be run once. 
Run the function on all subdirectories except the gating subdirectory (subdirectory we will use for gating, 
which is the first directory). Only needs to be run once to create the experimental details .csv files. 
The samplesheet should be sitting in the working directory, not in the subdirectories. 
```{r}
map(folders[-1], make_exp_details, samplesheet = "EE_GAP1_ArchMuts_2021.csv")
```

Next, copy the experiment details files from timepoints 1, 2, 4 and place them in the gating subdirectory. 
Only needs to be run once. 
On Mac, you can do this in the terminal or manually using your file GUI/Finder.
On PC, ??? 
```{bash}
cd ~/Google\ Drive/My\ Drive/greshamlab/projects/EE_GAP1_ArchMuts_Summer2021/data/vignette_data_workingcopy

scp ./01_EE_GAP1_ArchMuts_2021_061621_g8_GA/01_EE_GAP1_ArchMuts_2021_experiment_details.csv 01_02_04_EE_GAP1_ArchMuts_2021/ 

scp ./02_EE_GAP1_ArchMuts_2021_062121_g21_TD/02_EE_GAP1_ArchMuts_2021_experiment_details.csv 01_02_04_EE_GAP1_ArchMuts_2021/ 

scp ./04_EE_GAP1_ArchMuts_2021_062521_g37_IS/04_EE_GAP1_ArchMuts_2021_experiment_details.csv 01_02_04_EE_GAP1_ArchMuts_2021/
```

## STEP 2: Read in all files in the *gating subdirectory* and rename the channels.
Results in one timepoint's gating set containing all .fcs files, associated experiment details, and marker details

In our case, the first subdirectory is our gating subdirectory. 
```{r}
gating_subdir = folders[1]
```

```{r}
exp_details_path = list.files(path = paste0(gating_subdir), pattern = "_experiment_details.csv", full.names = T)
```

Edit Markers on Viewer pane. 
Type in B2-A in the marker column of the B2-A channel. FSC and SSC are included markers by default 
so we don't need to edit them. Click 'Save & Close' button at the bottom of the Viewer pane.
```{r}
timepoint_gating_set <- cyto_setup(path = paste0(gating_subdir), restrict=TRUE, select="fcs", details=F)
```

Add the experiment details file to the gating set
```{r}
experiment_details <- read_csv(exp_details_path, show_col_types = FALSE)
for(i in 1:length(names(experiment_details))){
  flowWorkspace::pData(timepoint_gating_set)[names(experiment_details[i])]<-experiment_details[i]
}
```

Check that the experiment details were successfully attached to the gating set. 
```{r}
cyto_details(timepoint_gating_set) %>% View()
```

By default the Experiment-Markers.csv file is named with the date created. 
Rename the experiment-markers.csv file. 
Need to do only once.
```{r}
file.rename(dir(pattern = "Experiment-Markers.csv"),"EE_GAP1_ArchMuts_2021-Experiment-Markers.csv")
```

## STEP 3:  Perform gating on gating set
Gate for 1) Cells, 2) Singlets, 3) CNVS
Results in a gating file, and gates applied to all samples in the gating set.

**Transform the data**
Transforming the data makes it easier to interpret visually and well as to draw gates. 
The untransformed data when plotted, the zero copy strains take up most of the coordinate plane 
while the one copy and two copy are very close together. The fluorescence profiles/distributions 
of the one copy and two copy controls overlap some. They are not mutually exclusive 
which is a limitation to note. 

This is a useful article if you I want think about to choosing different transformations:
https://dillonhammill.github.io/CytoExploreR/articles/CytoExploreR-Transformations.html

In our case, it was necessary to use a logical transformation for the B2-A values and 
a second transformation log for the other values. I could not use logical transformation
for all nor could I use the log transformation for all values. 

**IMPORTANT**: Run all of STEP 3 code in the Console, not inside the R notebook. Copy and paste it into the Console.
Make sure you set your working directory in the Console. 
```{r}
setwd("/Volumes/GoogleDrive/My Drive/greshamlab/projects/EE_GAP1_ArchMuts_Summer2021/data/vignette_data_workingcopy")
GFP_trans <- cyto_transformer_logicle(timepoint_gating_set,
                                      channels = c("B2-A"),
                                      widthBasis = -10
)#returns it as a list
FSC_SSC_trans <- cyto_transformer_log(timepoint_gating_set,
                                      channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H")
) #log transform the forward and side scatter

combined_trans <- cyto_transformer_combine(GFP_trans,FSC_SSC_trans) #combine both transformations

transformed_timepoint_gating_set <- cyto_transform(timepoint_gating_set,
                                                   trans = combined_trans) #applies the the transformation and returns it as a gatingSet
```

Check the transformed data by plotting
Subset rows to plot as needed.
Run this code in the console.
```{r}
cyto_plot_explore(transformed_timepoint_gating_set[c(2,14,16,17,18)], #choose rows to plot 
                  channels_x = "FSC-A",
                  channels_y = "B2-A",
                  axes_limits = "data")
```


#### Gating using the timepoints from the gating dataset.

*Note*:if you already have a gating template and don't need to draw gates,
skip the cyto_gate_draw() steps. Instead, use cyto_gatingTemplate_apply() 
to apply a gatingTemplate.csv to your gating set. 
See STEP 5.5.  

**IMPORTANT**: Run this in the Console not in the R Notebook 

A new window will pop out. Draw the desired gate.
Press esc when finished drawing to save the gate. 

First we gate for the cells.
```{r}
cyto_gate_draw(transformed_timepoint_gating_set,
  #transformed_logicle_timept,
               parent = "root",
               alias = "Cells",
               channels = c("FSC-A","SSC-A"),
               axes_limits = "data",
               gatingTemplate = paste0("cytek_gating_",version_name,".csv")
)
```

Then we define the singlets based on forward scatter height and width.
```{r}
cyto_gate_draw(transformed_timepoint_gating_set,
               parent = "Cells",
               alias = "Single_cells",
               channels = c("FSC-A","FSC-H"),
               axes_limits = "data",
               gatingTemplate = paste0("cytek_gating_",version_name,".csv")
)
```

Gating for CNVs using the 0,1 and 2 copy controls:
Subset the rows in the gatingset that you want to extract to overlay when drawing gates.
The row order correspond to the experimental details.  
Choose colors of the parent population and the overlay populations. 
Choose gate names

```{r}
experiment_details %>% View() #View the rows and choose which to extract
```

**IMPORTANT**: Run code in the Console not the R Notebook 
```{r}
DGY1 <- cyto_extract(transformed_timepoint_gating_set, "Single_cells")[c(5,11,17)] # we chose 3 rows of DGY1 data

DGY500 <- cyto_extract(transformed_timepoint_gating_set, "Single_cells")[c(1,7,13)] #we chose 3 rows of DGY500 data

DGY1315 <- cyto_extract(transformed_timepoint_gating_set, "Single_cells")[c(6,12,18)] #we chose 3 rows of DGY1315 data

cyto_gate_draw(transformed_timepoint_gating_set,
               parent = "Single_cells", #maps to first point color
               alias = c("zero_copy", "one_copy", "two_or_more_copy"), #Name the gate names here
               channels = c("FSC-A","B2-A"),
               axes_limits = "data",
               gatingTemplate = paste0("cytek_gating_",version_name,".csv"),
               overlay = c(DGY1, DGY500, DGY1315), #maps to remaining point colors
               point_col = c("gray", "green", "red", "blue") #choose colors
)
```

## STEP 4: Export gate-based data and non-gate-based single cell data tables with fluorescence values

This code can be run in the notebook now. 

Get cell count from each gate
```{r}
gs_pop_get_stats(transformed_timepoint_gating_set, c("Single_cells", "zero_copy", "one_copy", "two_or_more_copy")) %>%
    rename(Gate = pop, name = sample, Count = count) %>%
    left_join(experiment_details) %>%
    write_csv(paste0(version_name,"_counts.csv"))
```  

Get frequency of cells inside each gate
```{r}
  gs_pop_get_stats(transformed_timepoint_gating_set, c("Single_cells","zero_copy", "one_copy", "two_or_more_copy"), type = "percent") %>%
    rename(Gate = pop, name = sample, Frequency = percent) %>%
    left_join(experiment_details) %>%
    write_csv(paste0(version_name,"_freq.csv"))
```

Get single cell raw fluorescence data (NOT based on gates)
Compute the normalized B2-A fluoresence over cell size forward scatter area (FSC-A)
```{r}
timepoint_raw_list <- cyto_extract(transformed_timepoint_gating_set, parent = "Single_cells", raw = TRUE, channels = c("FSC-A", "B2-A")) #raw flow data of each single cell as a list of matrices

map_df(timepoint_raw_list, ~as.data.frame(.x), .id="name") %>% #convert to df, put list name in new column
  mutate(name = as.factor(name)) %>%  #convert `name` to factor
  left_join(experiment_details) %>% #join by name column to add metadata
  mutate(B2A_FSC = `B2-A`/`FSC-A`) %>%  #compute normalized fluorescence
  write_csv(paste0(version_name,"_SingleCellDistributions.csv"))
```

## STEP 5:  Use function to perform analysis to all subdirectories
This function that will
1. Read in all the files in a folder
2. Read in experiment details files using pData
3. Specify experiment markers
4. Transform gating set
5. Apply existing gating file using cyto_gatingTemplate_apply()
6. Output stats file as .csv

Specify markers and channels
```{r}
my_markers<-c("GFP") #list your marker name(s)
channel<-c("B2-A") #list your channel(s)
names(my_markers)<-channel
```

Function to analyze all experimental subdirectories
```{r}
analyze_all_exp = function(folder_name, my_markers, out_name=version_name, gating_template="cytek_gating.csv") {

  path <- folder_name #gets relative path name for folder to be analyzed

  prefix <- folder_name %>% str_extract("([0-9])+_EE_GAP1_ArchMuts_2021") #extracts the time point number from folder name

  exp_details_path <- paste0(path,"/",prefix,"_experiment_details.csv") #gets experiment details .csv from correct directory

  #1. read in files and make a gating set
  print(path)
  timepoint_gating_set <- cyto_setup(path=path, select="fcs", details=F, markers = F)

  #2. read in experiment details for that gating set
  experiment_details <- read_csv(exp_details_path, show_col_types = F) #import experiment-details.csv
  for(i in 1:length(names(experiment_details))){
    flowWorkspace::pData(timepoint_gating_set)[names(experiment_details[i])]<-experiment_details[i]
  }
  
  #3. specify markers for that gating set
  markernames(timepoint_gating_set)<-my_markers

  #4. transform data
  GFP_trans <- cyto_transformer_logicle(timepoint_gating_set,
                                                 channels = c("B2-A"),
                                                 widthBasis = -10
                                                 )#returns it as a list
  FSC_SSC_trans <- cyto_transformer_log(timepoint_gating_set,
                        channels = c("FSC-A", "FSC-H", "SSC-A", "SSC-H")
                        )
  combined_trans <- cyto_transformer_combine(GFP_trans,FSC_SSC_trans)
  transformed_timepoint_gating_set <- cyto_transform(timepoint_gating_set,
                                       trans = combined_trans) #applies the the transformation and returns it as a gatingSet

  #5. apply gatingTemplate.csv to transformed gating set
  cyto_gatingTemplate_apply(transformed_timepoint_gating_set, gatingTemplate= gating_template)

  #6. write data tables: count and frequency of cells inside each gate, single cell raw fluorescence

  #get cell count from each gate
  gs_pop_get_stats(transformed_timepoint_gating_set, c("Single_cells", "zero_copy", "one_copy", "two_or_more_copy")) %>%
    rename(Gate = pop, name = sample, Count = count) %>%
    left_join(experiment_details) %>%
    write_csv(paste0(out_name,"_counts_", prefix,".csv"))
  
  #get frequency of cells inside each gate
  gs_pop_get_stats(transformed_timepoint_gating_set, c("Single_cells","zero_copy", "one_copy", "two_or_more_copy"), type = "percent") %>%
    rename(Gate = pop, name = sample, Frequency = percent) %>%
    left_join(experiment_details) %>%
    write_csv(paste0(out_name,"_freq_", prefix,".csv"))

  #raw transformed B2-A and FSC values for each cell
  timepoint_raw_list <- cyto_extract(transformed_timepoint_gating_set, parent = "Single_cells", raw = TRUE, channels = c("FSC-A", "B2-A")) #raw flow data of each cell as a list of matrices
  map_df(timepoint_raw_list, ~as.data.frame(.x), .id="name") %>% #convert to df, put list name in new column
   mutate(name = as.factor(name)) %>% #convert `name` to factor
   left_join(experiment_details %>% #join by name column to add metadata
               mutate(generation = as.factor(unique(experiment_details$generation)))) %>% #convert generation to a factor
   mutate(B2A_FSC = `B2-A`/`FSC-A`) %>% #compute normalized fluor for each cell
   write_csv(paste0(out_name,"_SingleCellDistributions_",prefix,".csv"))
}
```

## STEP 6:  Apply function from STEP 5 to all subdirectories
Uses map() from purr to apply function analyze_all_exp() from step 5 to specified directories
Specify subdirectories and the gating template below. 

**IMPORTANT**: Run code in the Console, not the R Notebook
```{r}
map(folders[2:length(folders)],analyze_all_exp, my_markers, gating_template = paste0("cytek_gating_",version_name,".csv"))
```

## STEP 7: Pull in all counts or freq or single cell distribution files from directory and combine into a single dataframe

```{r}
list.files(path = ".", pattern = paste0(version_name,"_counts_([0-9])+_EE_GAP1_ArchMuts_2021")) %>%
  read_csv() %>%
  write_csv(file = paste0(version_name,"_counts_all_timepoints.csv"))

list.files(path = ".", pattern = paste0(version_name,"_freq_([0-9])+_EE_GAP1_ArchMuts_2021")) %>%
  read_csv() %>%
  write_csv(file = paste0(version_name,"_freq_all_timepoints.csv"))
```

For the Single Cell Data, do it on the HPC because large files. 
```{r}
list.files(path = ".", pattern = "01_02_04_v2_SingleCellDistributions") %>%
  read_csv() %>%
  write_csv(file = paste0(version_name,"_SingleCellDistributions_all_timepoints.csv")

```

Read in frequency csv, cell numbers csv's, single cell distributions for all timepoints
```{r}
freq = read_csv(paste0(version_name,"_freq_all_timepoints.csv"))

counts= read_csv(paste0(version_name,"_counts_all_timepoints.csv"))

sc_distr_alltimepoints <-read.csv(paste0(version_name,"_SingleCellDistributions_all_timepoints.csv"), 
  stringsAsFactors = T) %>% 
  mutate(generation = factor(generation, levels = unique(generation)))
```


Optional code to run on the HPC if you want to write subsetted single cell files for each population. 
```{r}
for(pop in unique(sc_distr_alltimepoints$sample)) {
  print(pop)
  sc_distr_alltimepoints %>%
  filter(sample == pop) %>%
  write_csv(paste0("sc_distributions_",pop,"_all_timepoints.csv"))
}
```

## STEP 8: Plot histograms ridgeplots using the single cell data

Examine the the ENTIRE DISTRIBUTION of B2-A (GFP) of the cells per population per generation.

First, plot ridgeplots of controls over time

Run STEP 8 code on the HPC because very large files.

Subset and write the control data files once.
Otherwise, read in the control data files. 
```{r}
#zero <- sc_distr_alltimepoints %>%
  #mutate(generation = factor(generation, levels = unique(generation))) %>%
  #filter(Description == "0 copy control") %>%
  #write_csv(file = "sc_distributions_0copyControl_all_timepoints.csv") #only need to run once
zero = read.csv("sc_distributions_0copyControl_all_timepoints.csv", stringsAsFactors = T) %>%
  mutate(generation = factor(generation, levels = unique(generation))) #convert generation to factor

#one <- sc_distr_alltimepoints %>% filter(Description == "1 copy control") %>%
  #write_csv(file = "sc_distributions_1copyControl_all_timepoints.csv") #only need to do once
one = read.csv("sc_distributions_1copyControl_all_timepoints.csv", stringsAsFactors = T) %>%
    mutate(generation = factor(generation, levels = unique(generation)))

#two <- sc_distr_alltimepoints %>% filter(Description == "2 copy control") %>%
  # write_csv(file = "sc_distributions_2copyControl_all_timepoints.csv") #only need do once
two = read.csv("sc_distributions_2copyControl_all_timepoints.csv", stringsAsFactors = T) %>%
    mutate(generation = factor(generation, levels = unique(generation)))

controls = bind_rows(zero, one, two)
```

Plot all controls in one ggplot, and facet by Description
```{r}
controls %>%
  filter(!(Description == "1 copy control" & generation == 203 |
         Description == "0 copy control" & generation == 231)) %>%
  ggplot(aes(B2A_FSC, generation, fill = Description)) +
  geom_density_ridges(scale = 1) +
  facet_grid(~Description) +
  scale_y_discrete(expand = expansion(add = c(0.2, 1.0)))+
  #scale_fill_brewer(type = "seq", palette = 5, direction = 1) +
  scale_fill_manual(values=c(RColorBrewer::brewer.pal(4, "Greens")[-1])) +
  scale_x_continuous("normalized fluorescence", limits=c(0, 2.5), breaks = c(0, 1, 2, 2.5), labels = c(0,1,2,2.5)) +
  theme_classic() +
  theme(
      legend.position = 'none', #remove the legend
      axis.text.x = element_text(family="Arial", size = 10, color = "black"), #edit x-tick labels
      axis.text.y = element_text(family="Arial", size = 10, color = "black"),
      strip.background = element_blank(), #remove box around facet title
      strip.text = element_text(size=12)
  )
ggsave("controls_generation_ridgeplot_facet.png")
```



## STEP 9: Assess gates

Make a data frame of frequency and counts
```{r}
freq_and_counts =
  counts %>% filter(Gate == "Single_cells") %>%
  rename(Parent = Gate) %>%
  left_join(freq) %>%
  filter(!(Gate == "Single_cells")) %>%
  mutate(Frequency = Frequency*100) %>%
  relocate(2:3, .after = Gate)
```

Make table of low cell (less than 70,000) observations. 
We will useful to have in order to anti_join() in further steps
```{r}
lowcell = freq_and_counts %>%
  filter(Count <7000) %>%
  mutate(generation = factor(generation, levels = unique(generation))) %>%
  select(-Count)
```

Check controls are in their proper gates
Adjust thresholds of pass/fail as desired. 
```{r}
fails = freq_and_counts %>%
  filter(Count>70000) %>% # exclude any well/timepoint with less than 70,000 single cells
  filter(str_detect(Description, "control")) %>%
  select(Description, Strain, generation, Gate, Frequency, name, Count) %>%
  mutate(flag = case_when(Strain == "DGY1" & Gate == "zero_copy" & Frequency >= 95 ~ "pass",
                          Strain == "DGY1" & Gate == "zero_copy" & Frequency < 95 ~ "fail",
                          Strain == "DGY1" & Gate == "one_copy" & Frequency >= 10 ~ "fail",
                          Strain == "DGY1" & Gate == "two_or_more_copy" & Frequency >=11 ~ "fail",
                          Strain == "DGY500" & Gate == "one_copy" & Frequency >= 79 ~ "pass",
                          Strain == "DGY500" & Gate == "one_copy" & Frequency < 79 ~ "fail",
                          Strain == "DGY500" & Gate == "zero_copy" & Frequency >= 11 ~ "fail",
                          Strain == "DGY500" & Gate == "two_or_more_copy" & Frequency >= 11 ~ "fail",
                          Strain == "DGY1315" & Gate == "two_or_more_copy" & Frequency >= 79 ~ "pass",
                          Strain == "DGY1315" & Gate == "two_or_more_copy" & Frequency < 79 ~ "fail",
                          Strain == "DGY1315" & Gate == "zero_copy" & Frequency >= 11 ~ "fail",
                          Strain == "DGY1315" & Gate == "one_copy" & Frequency >= 11 ~ "fail"
                          ))%>%
  filter(flag == "fail") %>%
  arrange(Description)

fails %>% write_csv(paste0(version_name,"_fails.csv"))
View(fails)
```


Plot controls over time
```{r}
freq_and_counts %>%
  filter(Count>70000,
          str_detect(Description, "control")) %>%
  select(Type, Strain, Description, generation, Gate, Frequency, Count) %>%
  anti_join(fails) %>% #exclude the the failed control timepoints which are likely contaminated
  ggplot(aes(generation, Frequency, color = Gate)) +
  geom_line() +
  facet_wrap(~Description) +
  ylab("% of cells in gate") +
  theme_minimal() +
  scale_x_continuous(breaks=seq(0,250,50)) +
  theme(text = element_text(size=12))
```


## STEP 10: Plot gate-based time series

Plot proportion of population in each gate over time for all experimental samples
```{r}
plot_list = list()
i=1
for(exp in unique(freq_and_counts$Description)) {
  #print(plot_dist(obs))
  plot_list[[i]] = freq_and_counts %>%
    filter(Count>70000,
           Description==exp) %>%
    ggplot(aes(generation, Frequency, color = Gate)) +
    geom_line() +
    facet_wrap(~sample) +
    ylab("% of cells in gate") +
    theme_minimal()+
    scale_x_continuous(breaks=seq(0,250,50))+
    theme(text = element_text(size=12))
  i = i+1
}
names(plot_list) = unique(freq_and_counts$Description)
plot_list$`GAP1 WT architecture` # change index to view replicates for different genetic backgrounds
plot_list$`GAP1 ARS KO`
plot_list$`GAP1 LTR KO`
plot_list$`GAP1 LTR + ARS KO`

```

Plot proportion of the population with a CNV over time
```{r}
freq_and_counts %>%
  filter(Count>70000) %>%
  filter(Gate %in% c("two_or_more_copy"), Type == "Experimental") %>%
  anti_join(fails)  %>% 
  group_by(sample, generation) %>%
  mutate(prop_CNV = sum(Frequency)) %>% View()
  select(sample, generation, Description, prop_CNV) %>%
  distinct() %>%
  ggplot(aes(generation, prop_CNV, color = sample)) +
  geom_line() +
  #geom_point()+
  facet_wrap(~Description) +
  xlab("Generation") +
  ylab("Proportion of the population with GAP1 CNV") +
  scale_color_manual(values = c("#DEBD52","#DBB741","#D7B02F","#CAA426","#D9BB59", #WT,5,gold
"#637EE7","#6F88E9","#7B92EA","#4463E2","#3053DF","#2246D7","#1E3FC3","#5766E6", #ALL,8,bluepurple
"#DE54B9","#E160BE","#E36CC3","#E578C8","#E885CD","#DB41B2","#D72FAA", #ARS,7,pink
"#54DE79","#41DB6A","#2FD75C","#26CA52","#23B84B","#60E182","#6CE38C","#78E595" #LTR,8,green
   )) +
  #scale_color_manual(values = c()
  theme_classic() +
  scale_x_continuous(breaks=seq(0,250,50)) +
  theme(text = element_text(size=12),
        legend.position = "none",
        axis.text.x = element_text(family="Arial", size = 10, color = "black"), #edit x-tick labels
        axis.text.y = element_text(family="Arial", size = 10, color = "black"),
        strip.background = element_blank(), #removed box around facet title
        strip.text = element_text(size=12)
        )
```


## STEP 11: Plot other non-gate-based time series 

Plot normalized median mCitrine fluorescence over time.
Overlay 0,1,2 controls on same graph as experimental with gray dotted lines.

on HPC, work with the single cell dataset and write the median normalized fluorescence data file.
```{r}
# on hpc, do once
sc_distr_alltimepoints %>%
  group_by(sample, generation) %>%
  mutate(Med_B2A_FSC = median(B2A_FSC)) %>%
  distinct(Med_B2A_FSC, .keep_all = T) %>%
  select(-FSC.A, -B2.A, -B2A_FSC) %>%
  write_csv("medians_normalized_fluor_alltimepoints.csv")
```
  
Back in the R notebook, do the remaining analysis  
```{r}
norm_medians = read_csv("medians_normalized_fluor_alltimepoints.csv") %>%
  left_join(cell_numbers) %>%
  select(-Marker) %>%
  rename(Gate = Population)
```

Rename description of controls so we can graph them on experimental facet plots  
```{r}
relabel_controls = norm_medians %>% 
  arrange(Description) %>%
  slice(rep(1:sum(str_detect(norm_medians$Type, "ctrl")), each = 4)) %>% #repeat each control row 4 times because was have 4 facetplots
  mutate(Description = rep(c("GAP1 ARS KO", "GAP1 LTR + ARS KO", "GAP1 LTR KO","GAP1 WT architecture"),    
  times=sum(str_detect(norm_medians$Type, "ctrl"))))
```

merge back to experimental rows from norm_medians df
```{r}
adj_norm_medians = merge(norm_medians %>% filter(Type == "Experimental"), relabel_controls, all = TRUE) %>% #merge back to experimental rows
  arrange(Description, generation) %>%
  filter(Count>70000) #exclude observations with <70,000 cells
```

Clean up data frame. Remove timepoints of controls that are abnormal as informed by ridgeplots
```{r}
clean_adj_norm_medians = adj_norm_medians %>%
  #remove select controls timepoints based on ridgeplots
  anti_join(adj_norm_medians %>% filter(generation == 116 & Type == "2_copy_ctrl")) %>%
  anti_join(adj_norm_medians %>% filter(generation == 182 & Type == "1_copy_ctrl")) %>%
  anti_join(adj_norm_medians %>% filter(generation == 203 & Type == "1_copy_ctrl")) %>%
  anti_join(adj_norm_medians %>% filter(generation == 231 & Type == "0_copy_ctrl")) %>%
  anti_join(adj_norm_medians %>% filter(generation == 260 & Type == "1_copy_ctrl"))
```

Graph experimental with along controls
```{r}
clean_adj_norm_medians %>%
  filter(!(Med_B2A_FSC<1.5 & Type == "Experimental")) %>%#filter out outliers (likely resulting from contamination) as defined by Fluor <1.5
ggplot(aes(generation, Med_B2A_FSC, color= sample)) +
  geom_line(aes(linetype = Type)) +
  scale_linetype_manual(values = c("dashed", "dashed", "dashed", "solid")) +
  scale_color_manual(values=c("gray", "gray", "gray", #controls
                              "#DEBD52", "#DBB741", "#D7B02F", "#CAA426","#D9BB59", #WT,5, golds
                              #rep("#5474DE", 8),  #ALL,8, blue/purple "#5474DE"
                              "#5774E5","#637EE7", "#6F88E9","#7B92EA","#4463E2","#3053DF","#2246D7","#1E3FC3", #ALL,8, blue/purple "#5474DE"
                              "#DE54B9","#E160BE","#E36CC3","#E578C8","#E885CD","#DB41B2","#D72FAA", #rep("#DE54B9", 7), #ARS, 7,  pink "#DE54B9"
                              "#54DE79","#41DB6A","#2FD75C","#26CA52","#23B84B","#60E182","#6CE38C","#78E595" #rep("#54DE79", 8)  #LTR,8, green "#54DE79"
  ))+
  facet_wrap(~Description) +
  xlab("Generation") +
  ylab("Median normalized fluorescence (a.u.)") +
  scale_x_continuous(breaks=seq(0,260,50)) +
  #ylim(c(1.5,2.5))+
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size=12),
        strip.background = element_blank(), #removed box around facet title
        strip.text = element_text(size=12),
        axis.text.x = element_text(family="Arial", size = 12, color = "black"), #edit x-tick labels
        axis.text.y = element_text(family="Arial", size = 12, color = "black"))
ggsave("MedNormFluo_FacetPlots.png")
```
